Performance analysis of high-resolution \\ ice sheet simulations

### 2022-10-07 paper/perfmod.tex
it is hard work to describe what Ua does

It is in this context that we can distinguish time-stepping types in coupled geometry-velocity, i.e.~mass and momentum conserving, glacier and ice sheet models.  A coupled model computes (\emph{fully}) \emph{implicit} time steps if no significant aspects of the geometry or velocity are held fixed during the (coupled) solution.  That is, in an implicit scheme a coupled and well-posed model for the updated ice surface elevation (or thickness) and velocity fields is solved.  A scheme which does not allow the map-plane ice-covered region to change during a time step cannot respond physically (conservatively) to a change in climate inputs; such a scheme is not actually implicit for the coupled problem and it cannot validly compute long-duration time steps.  Thus we say that a coupled scheme is \emph{semi-implicit} if, in particular, aspects of the ice geometry are held fixed during its coupled geometry-velocity solution.  A scheme is only semi-implicit if it computes margin advance or retreat after the velocity field is updated.  Finally, a scheme is \emph{explicit} for the coupled problem, and uses a time-step \emph{splitting} scheme \citep{HundsdorferVerwer2003}, if the ice geometry is held fixed during the velocity solution, and then this (accepted) velocity solution is used to update the geometry, e.g.~through an advection step for Equation \eqref{eq:masscontinuity}.

@@

The only currently implemented, unconditionally-stable, fully-implicit geometric-update solvers use the SIA stress balance.  For simplified (dome) geometry the scheme in \cite{Bueler2016}, based on Newton steps solved via direct linear algebra and single-grid ice margin determination, shows $\beta=0.8$.  (An earlier non-Newton implementation by \cite{JouvetBueler2012} scales worse.)  Convergence is robust for time steps of years to centuries on kilometer-scale grids for the Greenland ice sheet, using realistic, thus irregular, bed topographies.  Practical steady-state ($\Delta t=\infty$) solutions at ice sheet scale are also demonstrated by \cite{JouvetBueler2012} and \cite{Bueler2016}.

The SIA portions of hybrid time-stepping schemes by \cite{JouvetGraeser2013} and \cite{BrinkerhoffJohnson2015} are also solved implicitly.  \cite{JouvetGraeser2013} use a multigrid method, but their published, simplified-geometry results do not constrain $\beta$.  Overall time-stepping in these hybrids is only semi-implicit because the membrane-stress-resolving portion of the velocity explicitly advects the thickness, so the SIA-portion solvers are apparently only tested for time steps satisfying an advective condition $\Delta t < O(U^{-1}\Delta x)$ for $U$ which scales according to the sliding.
% equation (20) in JouvetGraeser2013
% equation (A21) in BrinkerhoffJohnson2015

For implicit Stokes time-stepping, a prospective, coupled, and free-boundary velocity and geometry-update solve is assumed to be, in the absence of constraining research, $O(m^{1+\gamma})$ for some $\gamma \ge \alpha$ to be determined.  One might also suppose $\gamma\ge \beta$, but in any case there are no implemented cases to measure.  These comments complete Table \ref{tab:performancemodel}, in which all estimates involve a scheme-dependent constant, something which is especially large for the implicit schemes.

% regarding the nMCD method proposed in [Bueler unpublished], which is expensive but multigrid, it may turn out that $\gamma$ is small

@@

From Table \ref{tab:performancemodel} we first observe a known property of explicit time-stepping for 2D (map-plane) diffusion equations such as SIA equation \eqref{eq:siamasscontinuity}, namely that effort, here flops per model year, scales as $O(\Delta x^{-4})$.  Recall that this follows because $\Delta t < O(\Delta x^2)$, and because the expense of one geometry-update operation is $O(m) = O(\Delta x^{-2})$.  Spatial mesh refinement by a factor of two therefore imposes an impressive 16-times increase in effort.

The \cite{Bueler2016} implicit SIA solver is not enough better, however.  Although long time steps can be taken by this implicit solver, the $\beta=0.8$ scaling gives $O(\Delta x^{-3.6})$ effort because $2+2\beta=3.6$.  However, an improvement to $\beta < 0.5$, presumably by application of a multigrid method, would transform such an implicit solver into a tool with notably superior performance compared to an explicit, $O(\Delta x^{-4})$ SIA scheme.

Now bypassing the over-simplified SIA stress balance, we see from Table \ref{tab:performancemodel} how the algorithmic scaling of solvers completely dominates.  In particular, an $\alpha=1$ explicit Stokes model, e.g.~one using direct, sparsity-exploiting linear algebra on each Newton step system, will do work which scales at the horrific rate $O(\Delta x^{-6})$ under a pessimistic stability condition.  Optimistic stability yields still-bad $O(\Delta x^{-5})$.  These simple observations emphasize the key role of multigrid-based stress balance solvers.  That is, even retaining explicit time-stepping, nearly-optimal ($\alpha \approx 0$) scaling of computation effort in Stokes and higher-order solvers will be necessary for routine application on high-resolution meshes.

On the other hand, suppose resolution ($\Delta x$) is fixed.  The Table also shows why algorithmic scaling remains important in the large-domain $L\to\infty$ limit when applying Stokes or higher-order dynamics.  An $\alpha=1$ method which might suffice for a smaller $L=100$ km ice cap will struggle for a $L=1000$ km ice sheet because the effort scales as the fourth ($2+2\alpha=4$) power of $L$.  By contrast, the computational work of a nearly-optimal solver will be proportional to ice sheet area $L^2$.

The promise of nearly-optimal solvers is profoundly revealed when implicit geometry updates are considered.  Future methods which simultanously satisfy the mass and momentum equations at each time step, and which do work essentially proportional to the number of degrees of freedom, have the greatest promise.  A $\gamma \approx 0$ implicit Stokes method would be hugely more capable for many tasks.  Specifically, and more achievably, a $\gamma < 0.5$ implicit, essentially unconditionally-stable, Stokes time-stepping method, presumably based on multigrid solution of the free-boundary problem for the coupled mass and momentum equations, is an appropriate goal for coming decades of research on numerical ice sheet models.  The computation cost would scale at $O(\Delta x^{-3})$, better than explicit SIA models, and the method would only update ice geometry when determined by the scientific need to resolve climate coupling, while avoiding shallow approximations.  The same goal makes sense for all membrane-stress-resolving solvers.  No apparent technical progress has yet been made on such an implicit Stokes design, and so these aspirations are decidedly long-term.  However, the above discussion suggests why measured values for $\alpha,\beta,\gamma$, or equivalent algorithmic scaling measures, are important performance metrics to report when describing new ice sheet solvers.

\subsubsection{Note added 4 July 2022.}  Errors in the first posted version (\url{https://arxiv.org/abs/2206.14352}) are corrected here.

>>>>

It is in this context that we can distinguish time-stepping types in coupled geometry-velocity, i.e.~mass and momentum conserving, glacier and ice sheet models.  A coupled model computes (\emph{fully}) \emph{implicit} time steps if no significant aspects of the geometry or velocity are held fixed during the (coupled) solution.  That is, in an implicit scheme a coupled and well-posed model for the updated ice surface elevation (or thickness) and velocity fields is solved.  A scheme which does not allow the map-plane ice-covered region to change during a time step cannot respond physically (conservatively) to a change in climate inputs; such a scheme is not actually implicit for the coupled problem and it cannot validly compute long-duration time steps.  Thus we say that a coupled scheme is \emph{semi-implicit} if, in particular, aspects of the ice geometry are held fixed during its coupled geometry-velocity solution, or if the velocity is held fixed during a time step as the geometry is updated.  A scheme which computes margin advance or retreat after the velocity field is updated is only semi-implicit, for example.  Finally, a scheme is \emph{explicit} for the coupled problem, and uses a time-step \emph{splitting} scheme \citep{HundsdorferVerwer2003}, if the ice geometry is held fixed during the velocity solution, and then this (accepted) velocity solution is used to update the geometry, e.g.~through an advection step for Equation \eqref{eq:masscontinuity}.

@@

The only currently implemented, unconditionally-stable, fully-implicit geometric-update solvers use the SIA stress balance.  For simplified (dome) geometry the scheme in \cite{Bueler2016}, based on Newton steps solved via direct linear algebra and single-grid ice margin determination, shows $\beta=0.8$.  (An earlier steady, Picard iteration implementation by \cite{JouvetBueler2012} scales worse.)  Convergence is also robust for implicit time steps of years to centuries on kilometer-scale grids for the Greenland ice sheet, using realistic, irregular, bed topographies.

Certain recent numerical models are semi-implicit in innovative ways.  The SIA portions of hybrid time-stepping schemes by \cite{JouvetGraeser2013} and \cite{BrinkerhoffJohnson2015} are solved implicitly, with nonnegativity of thickness in their SIA portions enforced as part of a free-boundary solution.  However, the time-stepping in these models is only semi-implicit (i.e.~time split) because the membrane-stress-resolving portion of the velocity explicitly advects thickness.  The SIA time-step solvers in these schemes have apparently only been tested for time steps satisfying an advective condition $\Delta t < O(U^{-1}\Delta x)$, for $U$ derived from sliding speeds.  Also, \cite{JouvetGraeser2013} use multigrid methods, but their published results do not constrain $\beta$.
% equation (20) in JouvetGraeser2013
% equation (A21) in BrinkerhoffJohnson2015
In a different design, the semi-implicit scheme in the open-source \'Ua numerical model \citep{Gudmundsson2021manual} implicitly solves the mass continuity equation, using an active set method, but while holding the membrane-stress-resolved velocity fixed.  Note that the formally third-order time-stepping in this model \citep{Gudmundsson2013}, while desirable for marine ice sheets, cannot achieve that accuracy when the nonnegativity-of-thickness constraint is actively determining the free boundary, e.g.~at retreating margins.
% equations (1), (26) in Gudmundson2013
% see semi-implicit comments in commit 9081fdbe3cf250d9cfd66504bf03134043ba2547
% https://github.com/GHilmarG/UaSource/blob/master/NexthTG3in2D.m: "Third order impicit Taylor-Galerkin (implict with respect to h, not u and v)"
% https://github.com/GHilmarG/UaSource/blob/master/SSS2dPrognostic.m: "semi-implicit forward integration. implicit with respect to thickness (h) explicit with respect to velocity"
% and "error('Semi-implicit time integration only implemented for SSTREAM and not SSHEET or Hybrid formulations \n')"

For implicit Stokes time-stepping, a coupled and free-boundary velocity and geometry-update solve is assumed to be, in the absence of constraining research, $O(m^{1+\gamma})$ for some $\gamma \ge \alpha$ to be determined.  One might also suppose $\gamma\ge \beta$, but there are no implemented cases to measure.  These comments complete Table \ref{tab:performancemodel}, in which all implicit scheme estimates involve an especially-large scheme-dependent constant.

@@

From Table \ref{tab:performancemodel} one first observes a known property of explicit time-stepping for 2D (map-plane) diffusion equations such as SIA equation \eqref{eq:siamasscontinuity}, namely that computational effort, here flops per model year, scales as $O(\Delta x^{-4})$.  Recall that this follows because $\Delta t < O(\Delta x^2)$, and because the expense of one geometry-update operation is $O(m) = O(\Delta x^{-2})$.  Spatial mesh refinement by a factor of two therefore imposes an impressive 16-times increase in effort.

The \cite{Bueler2016} implicit SIA solver is not enough better, however.  Although long time steps can be taken by this implicit solver, the $\beta=0.8$ scaling gives $O(\Delta x^{-3.6})$ effort.  However, an improvement to $\beta < 0.5$, presumably by application of a multigrid method, would transform such an implicit solver into a tool with notably superior performance compared to an explicit SIA scheme.

Now bypassing the over-simplified SIA stress balance, we see from Table \ref{tab:performancemodel} how the algorithmic scaling of solvers dominates performance concerns.  In particular, an $\alpha=1$ explicit Stokes model, e.g.~one using direct, sparsity-exploiting linear algebra on each Newton step system, will do work which scales at the horrific rate $O(\Delta x^{-6})$ under a pessimistic stability condition, and optimistic stability yields a still-bad $O(\Delta x^{-5})$ rate.  These simple observations emphasize the key role which will be played by multigrid-based stress balance solvers.  That is, even retaining explicit time-stepping, nearly-optimal ($\alpha \approx 0$) scaling of computation effort in Stokes and higher-order solvers will be necessary for routine application on high-resolution meshes.

On the other hand, suppose resolution ($\Delta x$) is fixed.  The Table also shows why algorithmic scaling remains important in the large-domain $L\to\infty$ limit when applying Stokes or higher-order dynamics.  The computational work of an optimal solver will be proportional to ice sheet area $L^2$.  However, an $\alpha=1$ method which might suffice for a smaller $L=100$ km ice cap will struggle for a $L=1000$ km ice sheet because the effort scales as the fourth ($2+2\alpha=4$) power of $L$.

The promise of nearly-optimal solvers is further revealed when implicit, coupled geometry-velocity updates are considered.  If they become possible, methods which simultanously satisfy the mass and momentum equations at each time step, by doing work (essentially) proportional to the number of degrees of freedom have the greatest promise.  A $\gamma \approx 0$ implicit Stokes method would be capable of many new tasks.  More achievably, a $\gamma < 0.5$ implicit, essentially unconditionally-stable, Stokes time-stepping method, presumably based on multigrid solution of the free-boundary problem for the coupled mass and momentum equations, is an appropriate goal for coming decades of research on numerical ice sheet models.  The computation cost would scale at $O(\Delta x^{-3})$, better than explicit SIA models, and the method would only update ice geometry when needed by ice-climate coupling, while avoiding shallow approximations.  (The same goal makes sense for any membrane-stress-resolving solver.)  No apparent technical progress has yet been made on such an implicit Stokes design, and these aspirations are decidedly long-term.  However, the above discussion suggests why measured values for $\alpha,\beta,\gamma$, or equivalent algorithmic scaling measures, are important performance metrics to report when describing new ice sheet solvers.
